---
title: "Figure 1 Gregor et al"
output:
  pdf_document: default
  html_notebook: default
---

The following is the code to create the figures in the paper "Mammalian gut metabolomes mirror microbiome composition and host phylogeny"

Load packages
```{r}

library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyr)
```

Load phyloseq objects and metadata

```{r}
#metadata
metadata <- read.delim("metadata.txt", header = T, sep = "") ## load metadata file
metadata <-subset(metadata, Type=="Feces")
metadata <- droplevels(metadata)


#16S sequencing results
ps_asv <- readRDS("Phyloseq_files/ps_asv.rds")

#LC-MS/MS metabolomics data (fecal samples, filtered to ~10k fecal peaks)
ps_pol <- readRDS("Phyloseq_files/ps_pol.rds")

#GC-MS metabolomics data (fecal samples)
ps_gcms <- readRDS("Phyloseq_files/ps_gcms.rds")

# PICRUSt data
ps_picrust <- readRDS("Phyloseq_files/ps_picrust.rds")

##Additional files
#LC-MS/MS metabolomics data, all samples and peaks
#ps_polar_all <- readRDS("Phyloseq_files/ps_polar_all.rds")
 
#GC-MS data, all samples
#ps_gcms_all <- readRDS("Phyloseq_files/ps_gcms_all.rds")
```



##Figure 1A- PCoAs

Create palettes for color (mammalian order) and shape (diet) (colors are from Okabe Ito scale)
```{r}
colorlist <- c(
  "#0072B2", #Artiodactyla
  "#CC79A7", #Carnivora
  "#009E73", #Perissodactyla
  "#F0E442", #Primates
  "#56B4E9" #Proboscidea
)

shapelist <- c(
  15, #Carnivores
  16,  # Herbivores
  17  #Omnivores
  )
```

PCoA analysis  
```{r}

#Choose a phyloseq object to work with: ps_gcms, ps_asv, or ps_pol
ps <- ps_asv

#Apply Bray Curtis ordination

ps.ord <- ordinate(ps, "PCoA", "bray")

#Or, apply alternate ordinations as shown in the SI
#ps.ord <- ordinate(ps, "PCoA", "euclidean")
#ps.ord <- ordinate(ps, "PCoA", "jaccard", binary=TRUE)
#ps.ord <- ordinate(ps, "PCoA", "wunifrac") ##16S only
#ps.ord <- ordinate(ps, "PCoA", "unifrac") ##16S only

##Examine eigenvalues
plot_ordination(ps, ps.ord, type= "scree")

##Create list of eigenvalues to normalize the axes
evals <- ps.ord$values[,1]
```
Make the plot object with ggplot2
```{r}
p <- plot_ordination(ps, ps.ord, color = "HostOrder", shape= "diet", title= "PCoA") 

p_final <- p + 
  coord_fixed(sqrt(evals[2] / evals[1])) +
  theme_bw(base_size=15)+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
  scale_color_manual(values=colorlist) +
  scale_shape_manual(values=shapelist)+
  geom_point(size=5)
p_final

```


##Figure 1B- Trees

Setup. This code closely follows examples laid out in the dendextend vignettes: https://cran.r-project.org/web/packages/dendextend/vignettes/dendextend.html#comparing-two-dendrograms

```{r}
library(ape)
library(dendextend)

```
For species with multiple individuals, we took the median values using dplyr to create a representative sample for the tree

```{r}

##Choose which phyloseq object to analyze (ps_gcms, ps_pol, ps_asv)
##eg 16S (ps_asv)
ps <- ps_asv

##The melting step is slow for large files, be patient
ps.m <- psmelt(ps)

ps.m <- ps.m %>% group_by(OTU, HostSpecies_taxonomy) %>% summarise_at(vars(Abundance), median, na.rm=TRUE)

#Reshape from long to wide format (unmelt)
m <- spread(ps.m, key= OTU, value= Abundance)

#give it rownames
m2 <- as.matrix(m[, -1])
rownames(m2) <- m$HostSpecies_taxonomy

##save median values as a new object: gcms_med, pol_med, asv_med
ps_med <- m2



```


Create the trees from the median data.

```{r}
##Cluster the median data
hclust.ps <- hclust(vegdist(ps_med, method = "bray"), method="ward.D2")

##Import host phylogeny
phylo_hosts <-read.tree(file="species.nwk")

##Convert to dendrograms for dendextend

dend_host <- as.dendrogram(phylo_hosts)
dend_ps <- as.dendrogram(hclust.ps)
```

Make each tanglegram by pairing dend_host with the tree from each dataset. Colored lines highlight subtrees present in both.
```{r}
##make a dendlist with the two dendrograms to compare: eg host phylogeny and 16S
dendlist <- dendlist(dend_host, dend_ps)

tanglegram(dendlist, sort=T, main_left = "Host phylogeny", main_right = "16S", common_subtrees_color_branches = FALSE, lwd=1, lab.cex= 1, margin_inner = 7, main="", columns_width= c(5,1,5), highlight_branches_lwd=FALSE, highlight_distinct_edges=FALSE)


#How is the entanglement (quality of alignment, lower is better)?
dendlist %>% entanglement

#Correlation

cor.dendlist(dendlist, method_coef = "pearson")
```

Since the observations creating the Baker's Gamma Index of such a measure are correlated, we need to perform a permutation test for the calculation of the statistical significance of the index. We will look at the distribution of Baker's Gamma Index under the null hypothesis (assuming fixed tree topologies). This will be different for different tree structures and sizes. Here are the results when the compared tree is compared to itself after shuffling its own labels (the_cor, correlation of 1), and when comparing tree 1 to the shuffled tree 2 (the_cor2):
```{r}
set.seed(42)
the_cor <- cor_bakers_gamma(dend_ps, dend_ps)
the_cor2 <- cor_bakers_gamma(dend_ps, dend_host)
the_cor
the_cor2
```

Create the null model.
```{r}
dend_mixed <- dend_ps

R <- 1000
cor_bakers_gamma_results <- numeric(R)


for(i in 1:R) {
   dend_mixed <- sample.dendrogram(dend_mixed, replace = FALSE)
   cor_bakers_gamma_results[i] <- cor_bakers_gamma(dend_host, dend_mixed)
}

##Create plot to show results
plot(density(cor_bakers_gamma_results),
     main = "16S to host phylogeny, Baker's gamma distribution under H0",
     xlim = c(-1,1))
abline(v = 0, lty = 2)
abline(v = the_cor, lty = 2, col = 2)
abline(v = the_cor2, lty = 2, col = 4)
legend("topleft", legend = c("16S to itself", "16S to host phylogeny"), fill = c(2,4))
round(sum(the_cor2 < cor_bakers_gamma_results)/ R, 4)

title(sub = paste("One sided p-value:",
                  "to itself =",  round(sum(the_cor < cor_bakers_gamma_results)/ R, 4),
                  " ; to host phylogeny tree =",  round(sum(the_cor2 < cor_bakers_gamma_results)/ R, 4)
                  ))
```


##Fig 1C- Comparison of beta dissimilarity for each dataset

Examine overall distances in different datasets 

```{r}
##Create function to compare distances, based on a function written by Jeff Kimbrel: https://jeffkimbrel.github.io/

allDistances = function(p = ps) {

dist = phyloseq::distance(p, method = "bray")
dist.m = melt(as.matrix(dist))
# remove self-comparisons
  dist.m = dist.m %>%
    filter(as.character(Var1) != as.character(Var2)) %>%
    mutate_if(is.factor,as.character)
  #make names of comparisons and sort
  dist.m$name <- paste0(dist.m[,1],"_",dist.m[,2])
   dist.m <-dist.m[order(dist.m[,4]),]

}
##Perform on all relevant datasets

dis_asv <- allDistances(p= ps_asv)
dis_pol <- allDistances(p= ps_pol)
dis_gcms <- allDistances(p= ps_gcms)
dis_pic <- allDistances(p= ps_picrust)

##Merge into single dataframe
  dis <- data.frame(dis_asv[,4], dis_asv[,1], dis_asv[,3], dis_pic[,3], dis_pol[,3], dis_gcms[,3])
  colnames(dis) <-c("pair", "animal_1", "Bacterial composition (16S)", "PICRUSt2", "LC-MS", "GC-MS")
  

##Melt dataframe
  dis.melt <- melt(dis, id= c("pair", "animal_1"))
  colnames(dis.melt) <- c("pair", "animal_1", "data_type", "Distance")
   
  
  
# Plot the data as boxplots. First use rev to reverse the order of the factor levels.
  dis.melt$data_type <- factor(dis.melt$data_type, levels=rev(levels(dis.melt$data_type)))
  ggplot(data=dis.melt, aes(x=data_type, y=Distance)) +
    theme_bw(base_size=15) +
    geom_boxplot(outlier.shape = NA, fill=NA, size=1) +
    labs(title= "Beta diversity by method", x = "", y = "Dissimilarity")+
    ylim(0,1)+
    coord_flip()

```
Calculate adjusted p-values using one-way ANOVA analysis followed by Tukey’s test
```{r}
# Compute the analysis of variance
res.aov <- aov(dis.melt$Distance ~ dis.melt$data_type)
# Summary of the analysis
summary(res.aov)
#Tukey's test
TukeyHSD(res.aov) 
```




##Mantel and Permanova analyses

PERMANOVA analysis

```{r}
set.seed(42)

##Pick a dataset: ps_asv, ps_gcms, ps_pol
ps <- ps_asv

# Calculate dissimilarity matrix
dis <- phyloseq::distance(ps, method = "bray")

# Adonis test
adonis(dis ~ diet+digestive_strategy+HostOrder+HostSpecies+collection_time, data = metadata)

```


For subsets analysis:

```{r}
set.seed(42)

##choose a dataset: ps_asv, ps_pol, ps_gcms
ps <- ps_asv
data <- as(otu_table(ps), "matrix")
```

Create one of the relevant subsets:
```{r}


#n=2 (change filters accordingly for n=3, 4, 5)
set.seed(42)
metadata2 <-bind_rows(
  data.frame(metadata) %>%
    group_by(HostSpecies) %>%
    filter(n()>1) %>%
    sample_n(size =2),
  data.frame(metadata) %>%
    group_by(HostSpecies) %>%
    filter(n()==1))
  
##Special cases for n=1:

#n=1
#set.seed(42)
#metadata2 <-    data.frame(metadata) %>%
  #group_by(HostSpecies) %>% 
  #sample_n(size =1) 


#Or, n=1 with median values
#ps.m <- ps.m %>% 
 # group_by(OTU, HostSpecies_taxonomy) %>% 
  #summarise_at(vars(Abundance), median, na.rm=TRUE)
```

Select the subset of samples from the data and run PERMANOVA test again

```{r}
##trim data to match subsetted samples
subset_n <- data[rownames(data) %in% metadata2$sample_name,]

##remove empty columns
subset_n <- subset_n[,colSums(subset_n)>0]
dim(subset_n)

#make distance matrix
dis <- vegdist(subset_n, method="bray")


adonis(dis ~ diet+digestive_strategy+HostOrder+HostSpecies+collection_time, data = metadata2)
```

Analysis for samples for which sex was recorded (Table S4)
```{r}
##Choose a dataset: ps_asv, ps_pol, ps_gcms

ps <- ps_pol

#first remove samples for which sex was not recorded
metadata2 <-subset(metadata, Sex=="F"|Sex=="M")

ps <- prune_samples(metadata2$sample_name, ps)

dis <- phyloseq::distance(ps, method = "bray")

adonis(dis ~ diet+digestive_strategy+HostOrder+HostSpecies+Sex+collection_time, data = metadata2)
```

Mantel correlations

```{r}

dis_pol <- phyloseq::distance(ps_pol, method = "bray")
dis_gcms <- phyloseq::distance(ps_gcms, method = "bray")
dis_asv <- phyloseq::distance(ps_asv, method = "bray")

#Note: check order of samples matches
#match(attr(dis_asv,"Labels"),attr(dis_gcms,"Labels"))
#match(attr(dis_asv,"Labels"),attr(dis_pol,"Labels"))

#Perform Mantel test to compare datasets
mantel(dis_asv, dis_gcms, method="pearson", permutations = 9999)

mantel(dis_asv, dis_pol, method="pearson", permutations = 9999)

mantel(dis_pol, dis_gcms, method="pearson", permutations = 9999)

##Function to subset data as in SI (or, use one of the subsets above)
#ps_asv_subset <- subset_samples(ps_asv, HostSpecies=="Rhino")



```

Or, for subsets, first generate metadata2 as in the PERMANOVA test above for the sample subset of interest 
```{r}
asv <- as(otu_table(ps_asv), "matrix")
gcms <- as(otu_table(ps_gcms), "matrix")
pol <- as(otu_table(ps_pol), "matrix")

##trim data accordingly
asv <- asv[rownames(asv) %in% metadata2$sample_name,]
pol <- pol[rownames(pol) %in% metadata2$sample_name,]
gcms <- gcms[rownames(gcms) %in% metadata2$sample_name,]

##remove empties
asv <- asv[,colSums(asv) >0]
pol <- pol[,colSums(pol) >0]
gcms <- gcms[,colSums(gcms) >0]

#Compare the datasets of interest, eg 16S and GC-MS
mantel(vegdist(asv), vegdist(gcms), method="pearson", permutations = 9999)
```




