---
title: "Data analysis Gregor et al"
output:
  pdf_document: default
  html_notebook: default
---

The following is the code to create the figures in the paper "Mammalian gut metabolomes mirror microbiome composition and host phylogeny"

Load packages
```{r}

library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyr)
```

Load phyloseq objects and metadata

```{r}
#metadata
metadata <- read.delim("metadata.txt", header = T, sep = "") ## load metadata file
metadata <-subset(metadata, Type=="Feces")
metadata <- droplevels(metadata)

##Short chain fatty acid data from GC-FID
ps_scfa <- readRDS("Phyloseq_files/ps_scfa.rds")


#GC-MS metabolomics data (fecal samples)
ps_gcms <- readRDS("Phyloseq_files/ps_gcms.rds")


```


##Fig 3A- SCFAs

Calculate ratios of acetate, propionate, and butyrate
```{r}

aceprobut <- as.data.frame(otu_table(ps_scfa))
aceprobut <- aceprobut[,c(1:2,4)]

#Normalize by sum
rowsums <- rowSums(aceprobut)
aceprobut <- aceprobut/rowsums
```


Make figure 3A
```{r}

#Order the samples like in heatmaps
aceprobut$sample_name <- rownames(aceprobut)

aceprobut.melt <- melt(aceprobut)

aceprobut.melt$sample_name <-factor(aceprobut.melt$sample_name, levels=c("Hyena_1","Jungle_cat_1","Jungle_cat_2","Sand_cat_1","Wolf_1","Wolf_2","Leopard_1","Leopard_2","Lion_1","Lion_2","Lion_3","Tiger_1","Tiger_2","Black_bear_1","Black_bear_2","Brown_bear_1","Brown_bear_2","Brown_bear_3","Brown_bear_4","Brown_bear_5","Coati_1","Coati_3","Coati_4","Coati_5","Coati_6","Coati_7","Coati_8","Af_elephant_1","Af_elephant_2","Af_elephant_3","As_elephant_1","As_elephant_2","As_elephant_3","As_elephant_4","As_elephant_5","Donkey_1","Zebra_1","Zebra_10","Zebra_11","Zebra_12","Zebra_13","Zebra_14","Zebra_15","Zebra_16","Zebra_2","Zebra_3","Zebra_4","Zebra_5","Zebra_6","Zebra_7","Zebra_8","Zebra_9","Rhino_1","Rhino_10","Rhino_11","Rhino_12","Rhino_13","Rhino_2","Rhino_3","Rhino_4","Rhino_5","Rhino_6","Rhino_7","Rhino_8","Rhino_9","Goat_1","Sheep_1","Sheep_2","Sheep_3","Sheep_4","Sheep_5","Gorilla_1","Gorilla_2","Gorilla_3","Gorilla_4","Gorilla_5","Gorilla_6","Chimpanzee_1","Chimpanzee_10","Chimpanzee_11","Chimpanzee_12","Chimpanzee_13","Chimpanzee_4","Chimpanzee_5","Chimpanzee_6","Chimpanzee_7","Chimpanzee_9","Mandarill_1","Mandarill_2","Mandarill_3","Mandarill_4","Mandarill_5","Mandarill_6","Mandarill_7","Mandarill_8","Capuchin_1","Gibbon_1","Gibbon_2","Gibbon_3","Lemur_1","Lemur_2"))

greys <-c("#f0f0f0",
"#636363",
"#bdbdbd")


##Make the bar plot
ggplot(aceprobut.melt, aes(x=value, y=sample_name)) +
  geom_col(aes(fill=variable))+
  scale_fill_manual(values =greys)

```
##Fig 3B-D- Boxplots of GCMS metabolites

Create palettes for color (mammalian order) and shape (diet) (colors are from Okabe Ito scale)
```{r}
colorlist <- c(
  
  "#CC79A7", #Carnivora
  "#F0E442", #Primates
   "#009E73", #Perissodactyla
  "#56B4E9", #Proboscidea
  "#0072B2" #Artiodactyla
)

shapelist <- c(
  15, #Carnivores
  16,  # Herbivores
  17  #Omnivores
  )
```

Prepare the data for visualization
```{r}

metadata$Order_diet <-factor(metadata$Order_diet, levels=c("Carnivora_Carnivores",  "Carnivora_Omnivores","Primates_Herbivores","Primates_Omnivores",      
"Perissodactyla_Herbivores","Proboscidea_Herbivores","Artiodactyla_Herbivores")) 

metadata$HostOrder <-factor(metadata$HostOrder, levels=c("Carnivora","Primates",      
"Perissodactyla","Proboscidea","Artiodactyla")) 

metab <- as(otu_table(ps_gcms), "matrix")
```

For lactate and succinate, we will use both the diet and the order on the x axis to show the split within Carnivora
```{r}
##Select the metabolite you want to visualize

#Lactate 36_RT4.35_BAL100
#Succinate 228_RT11.13_BAL100

##Example: lactate
metabolite <-metab[,"36_RT4.35_BAL100"]

ggplot(metadata, aes(x=Order_diet, y=metabolite, color=HostOrder)) +
  theme_bw(base_size=15)+
  geom_point(aes(col = HostOrder, shape=diet), size=2.5, position="jitter") +
  geom_boxplot(color = "black", outlier.shape=NA, fill=NA) + 
  labs(title= "", x = "", y = "") +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x=element_blank()) +
  scale_colour_manual(values = colorlist) +
  scale_shape_manual(values = shapelist)

```

Anova:

```{r}
# Compute the analysis of variance
res.aov <- aov(metabolite ~ metadata$Order_diet)
# Summary of the analysis
summary(res.aov)
```

If Anova is significant, perform Tukey multiple pairwise comparisons:

```{r}
TukeyHSD(res.aov)
```

For panels C and D, we will group by the order on the x axis
```{r}
#Cadaverine 531_RT22.1_BAL100
#5-aminovalerate 442_RT18.22_BAL95
#Putrescine 485_RT20.31_BAL100
#3-hydroxyphenylacetic acid 428_RT17.8_BAL100

##Example: cadaverine
metabolite <-metab[,"531_RT22.1_BAL100"]

ggplot(metadata, aes(x=HostOrder, y=metabolite, color=HostOrder)) +
  theme_bw(base_size=15)+
  geom_point(aes(col = HostOrder, shape=diet), size=2.5, position="jitter") +
  geom_boxplot(color = "black", outlier.shape=NA, fill=NA) + 
  labs(title= "", x = "", y = "") +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x=element_blank()) +
  scale_colour_manual(values = colorlist) +
  scale_shape_manual(values = shapelist)



```
Anova:

```{r}
# Compute the analysis of variance
res.aov <- aov(metabolite ~ metadata$HostOrder)
# Summary of the analysis
summary(res.aov)
```

If Anova is significant, perform Tukey multiple pairwise comparisons:

```{r}
TukeyHSD(res.aov)
```

Fig S16 and S17: PICRUSt2 analysis figures in the SI

```{r}
# PICRUSt data
ps_picrust <- readRDS("Phyloseq_files/ps_picrust.rds")

metab <- t(as(otu_table(ps_picrust), "matrix"))

#Fig S16:
#KEGG acrylate pathway enzyme K01026
#Fig S17:
# KEGG lysine decarboxylase K01582
# KEGG aminoaldehyde dehydrogenase K00137
# KEGG 5-aminovalerate aminotransferase K07250
#	KEGG ornithine to putrescine K01581

##Example: cadaverine
metabolite <-metab[,"K01026"]

##Then repeat above steps to make boxplots

```


##SCFA Boxplots in SI:

Fig S15A (by SCFA)
```{r}
#Melt the phyloseq object
ps_scfa.m <- psmelt(ps_scfa)

ps_scfa.m$HostOrder <-factor(ps_scfa.m$HostOrder, levels=c("Carnivora","Primates",      
"Perissodactyla","Proboscidea","Artiodactyla"))


ggplot(ps_scfa.m, aes(x=HostOrder, y=Abundance)) +
theme_bw(base_size=15)+
facet_wrap(vars(OTU), scales="free", nrow=2) +
geom_point(aes(col = HostOrder, shape = diet), size=2, position="jitter") +
geom_boxplot(color = "black", outlier.shape=NA, fill=NA) +
scale_colour_manual(values = colorlist) +
scale_shape_manual(values = shapelist) +
labs(title= "", x = "", y = "Concentration (mmol/kg feces)")+ 
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x=element_blank())

```
Fig S15B (by Order)
```{r}
#Subset dataset

ps_aceprobut.m <- subset(ps_scfa.m, OTU=="Acetate"|OTU=="Propionate"|OTU=="Butyrate")

ps_aceprobut.m$OTU <-factor(ps_aceprobut.m$OTU, levels=c("Acetate","Propionate",      
"Butyrate"))


ggplot(ps_aceprobut.m, aes(x=OTU, y=Abundance)) +
theme_bw(base_size=15)+
facet_wrap(vars(HostOrder, diet), scales="free", nrow=2) +
geom_point(aes(col = HostOrder, shape = diet), size=2, position="jitter") +
geom_boxplot(color = "black", outlier.shape=NA, fill=NA) +
scale_colour_manual(values = colorlist) +
scale_shape_manual(values = shapelist) +
labs(title= "", x = "", y = "Concentration (mmol/kg feces)")+ 
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
scale_x_discrete(labels=c("Acetate" = "Ace", "Propionate" = "Pro",
                              "Butyrate" = "But"))

```

















